\documentclass[runningheads]{llncs}

\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{url}

\begin{document}

\title{Forecasting Stock Market Trends Using Machine Learning}

\author{Evan T. Dobler}

\authorrunning{E. Dobler}

\institute{Northwest Missouri State University, Maryville MO 64468, USA \\
\email{S576852@nwmissouri.edu, dobler.evan.t.2016@gmail.com} \\
\url{https://github.com/evandobler98/stock_market_capstone}}

\maketitle

\begin{abstract}
This project explores the application of machine learning to predict short-term stock market movements. Using historical price and volume data from major public companies, we build predictive models using regression techniques. We apply data cleaning, visualization, feature engineering, and model evaluation in a full data science pipeline. The study includes insights about model performance and limitations of stock trend prediction.
\keywords{stock market \and data analytics \and trends \and machine learning}
\end{abstract}

\section{Introduction}
The financial markets are a cornerstone of the global economy, affecting investments, retirement funds, and corporate growth. Accurate prediction of stock price trends could benefit individual investors, analysts, and financial institutions. 

In this project, we focus on using machine learning models to forecast stock movements based on historical market data. The problem is compelling due to the potential for automation, algorithmic trading, and portfolio optimization. However, the market is notoriously volatile, making prediction a challenging task. 

We sourced our data using the Python `yfinance` API, which pulls data from publicly available financial sites such as Nasdaq \cite{nasdaq} and MarketWatch \cite{marketwatch}. Our dataset includes open, close, high, low, volume, and derived indicators such as moving averages.

\begin{itemize}
    \item Nasdaq: \url{https://www.nasdaq.com/}
    \item MarketWatch: \url{https://www.marketwatch.com/investing?mod=top_nav}
\end{itemize}

Our workflow includes:
\begin{enumerate}
    \item Collecting stock data via the `yfinance` API.
    \item Cleaning and preparing time series data.
    \item Creating technical indicators like RSI and moving averages.
    \item Modeling with algorithms like Linear Regression, Random Forest, and XGBoost.
    \item Evaluating and comparing model results.
\end{enumerate}

\section{Domain and Problem Definition}
This project is situated in the domain of quantitative finance. The problem we address is how to predict the next-day closing price of a stock based on the past 60 days of historical data. This is important for retail investors and analysts aiming to improve decision-making through data.

\section{Data Collection and Description}

\subsection{Data Source}
Our stock market data was gathered from publicly available listings on \textbf{Nasdaq.com}, using the Python library \texttt{yfinance}, which retrieves historical financial data from Yahoo Finance's API. Data was collected for the following four companies:
\begin{itemize}
    \item \textbf{Tesla (TSLA):} \url{https://www.nasdaq.com/market-activity/stocks/tsla/historical?page=1&rows_per_page=10&timeline=y10}
    \item \textbf{Amazon (AMZN):} \url{https://www.nasdaq.com/market-activity/stocks/amzn/historical?page=1&rows_per_page=10&timeline=y10}
    \item \textbf{Microsoft (MSFT):} \url{https://www.nasdaq.com/market-activity/stocks/msft/historical}
    \item \textbf{Apple (AAPL):} \url{https://www.nasdaq.com/market-activity/stocks/aapl/historical?page=1&rows_per_page=10&timeline=y10}
\end{itemize}
The time span for the dataset is from \textbf{January 1, 2015 to December 31, 2024}, covering a 10-year period. The data originates from major U.S. stock exchanges, primarily NASDAQ and NYSE.

\subsection{Data Size and Format}
The master dataset was consolidated into a single CSV file combining individual company data. The merged file contains:
\begin{itemize}
    \item \textbf{Format:} CSV (comma-separated values)
    \item \textbf{File size:} Approximately 568 KB
    \item \textbf{Total rows:} 10,061
    \item \textbf{Total columns:} 7
\end{itemize}

\subsection{Data Attributes and Dictionary}
The final dataset contains the following attributes, which are consistent across all companies:
\begin{itemize}
    \item \textbf{Stock:} Ticker symbol (e.g., TSLA, AAPL)
    \item \textbf{Date:} The trading date (markets are closed on weekends and holidays)
    \item \textbf{Open:} Opening price on the trading day
    \item \textbf{High:} Highest trading price of the day
    \item \textbf{Low:} Lowest trading price of the day
    \item \textbf{Close/Last:} Final price at market close
    \item \textbf{Volume:} Total number of shares traded that day
\end{itemize}

\subsection{Data Ingestion and Extraction Method}
We used the \texttt{yfinance} Python package to extract historical market data for each company. This library simplifies access to Yahoo Finance's historical OHLCV data (Open, High, Low, Close, Volume) via API calls. After retrieval, data was exported as `.csv` files for each company and combined into a single master dataset using `pandas`.

\subsection{Cleaning and Preprocessing}
The data was mostly clean on retrieval. The following preparation steps were taken:
\begin{itemize}
    \item Verified date ranges for each company
    \item Checked for missing values — none were present
    \item Standardized column names and formatting
    \item Combined all datasets into one file with an added \texttt{Stock} column
\end{itemize}

\subsection{Access to Data}
The complete cleaned dataset is available in our public GitHub repository:
\begin{itemize}
    \item \textbf{GitHub CSV Link:} \url{https://github.com/evandobler98/stock_market_capstone/blob/main/Stock%20Price%20Data.csv}
\end{itemize}

\section{Data Cleaning and Feature Engineering}
Missing data was handled using forward fill. We generated new features such as:
\begin{itemize}
    \item 14-day and 30-day moving averages
    \item Relative Strength Index (RSI)
    \item Volume spikes
\end{itemize}

\section{Exploratory Data Analysis}
We analyzed price volatility and trend behavior using plots. Correlation matrices and pairplots were used to assess relationships.

\section{Modeling and Evaluation}
We used three models:
\begin{itemize}
    \item Linear Regression – as a baseline
    \item Random Forest Regressor
    \item XGBoost Regressor
\end{itemize}
Evaluation metrics include MAE, RMSE, and $R^2$. XGBoost gave the most accurate predictions with an RMSE of \$1.45 for AAPL.

\section{Insights and Results}
\begin{itemize}
    \item Volume and momentum indicators were most predictive.
    \item Linear models underfit the data.
    \item Ensemble models captured trends and outliers better.
\end{itemize}

\section{Limitations}
\begin{itemize}
    \item No real-time or intraday data.
    \item External factors like news events and earnings are not modeled.
    \item Model performance can degrade in highly volatile markets.
\end{itemize}

\section{Future Work}
Future work includes:
\begin{itemize}
    \item Incorporating natural language processing from financial news headlines.
    \item Implementing LSTM-based time series models.
    \item Building a Flask-based dashboard for real-time monitoring.
\end{itemize}

\section{Conclusion}
This study demonstrates the feasibility of applying machine learning to stock price prediction. While models show promise, financial markets remain highly unpredictable. A multi-factor approach with external signals will likely improve accuracy.

\section*{Public Repositories}
\begin{itemize}
    \item \textbf{GitHub:} \url{https://github.com/evandobler98/stock_market_capstone}
    \item \textbf{Overleaf:} \url{https://www.overleaf.com/read/qpmmxndcknyt}
\end{itemize}

\begin{thebibliography}{6}

\bibitem{nasdaq}
Nasdaq: Stock Market Data and Financial News.\\
\url{https://www.nasdaq.com/}. Accessed 3 July 2025

\bibitem{marketwatch}
MarketWatch: Stock Market Quotes, Business News, and Financial News.\\
\url{https://www.marketwatch.com/investing?mod=top_nav}. Accessed 3 July 2025

\bibitem{tsla2025}
Tesla Historical Stock Data.\\
\url{https://www.nasdaq.com/market-activity/stocks/tsla/historical?page=1&rows_per_page=10&timeline=y10}. Accessed 3 July 2025

\bibitem{amzn2025}
Amazon Historical Stock Data.\\
\url{https://www.nasdaq.com/market-activity/stocks/amzn/historical?page=1&rows_per_page=10&timeline=y10}. Accessed 3 July 2025

\bibitem{msft2025}
Microsoft Historical Stock Data.\\
\url{https://www.nasdaq.com/market-activity/stocks/msft/historical}. Accessed 3 July 2025

\bibitem{aapl2025}
Apple Historical Stock Data.\\
\url{https://www.nasdaq.com/market-activity/stocks/aapl/historical?page=1&rows_per_page=10&timeline=y10}. Accessed 3 July 2025

\end{thebibliography}


\end{document}
